---
title: "2016 Nebraska Presidential Election Contributions"
author: "Nathaniel Booth"
date: "March 20, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(dplyr)
library(tidyr)
```

```{r echo=FALSE, load_data}
ne = read.csv('C:\\Users\\Nathan\\Desktop\\EDA\\ne_data.csv', 
                  row.names = NULL)
colnames(ne) <- c(colnames(ne)[-1], 'x')
```

# Introduction

This is Federal Election Commission [contribution data](http://fec.gov/disclosurep/PDownload.do) for the 2016 presidential election. It is limited to individual Nebraska contributors. This doesn't include contributions made to or on behalf of candidates by businesses or political action committees (like the infamous Super PACs). Thus, this is an incomplete picture of the flow of money in Nebraska during this election.

### Background

The maximum that an individual can contribute during a single election is $2,700. Primaries and general elections are separated. So, an individual can contribute up to the maximum amount for the primary and general elections each, to reach a maximum of $5,400.

The data includes both the general election and each party's primary elections. An explanation for each variable is [here](ftp://ftp.fec.gov/FEC/Presidential_Map/2016/DATA_DICTIONARIES/CONTRIBUTOR_FORMAT.txt).

# Univariate Exploration

```{r echo=FALSE, data_structure}
ne$cmte_id <- as.factor(ne$cmte_id)
ne$contb_receipt_dt <- as.Date(ne$contb_receipt_dt, "%d-%b-%y")
ne <- subset(ne, select = -c(contbr_st, x))
str(ne)
```

The dataset has mostly factor variables, but contributions are numbers. I removed the state variable because it is constant for all rows. I converted the committee ID from character to string. If I use it later, then I anticipate it to be more useful in this form. I converted the receipt date of a contribution to a date object. Maybe I'll find some interesting clustering of contributions in time.

#### Initial Questions and Thoughts
* Are certain employers overrepresented?
* Are certain occupations overrepresented?
* It might be useful to create a new variable that assigns party alignment for each candidate.
* Apparently, there were 22 different candidates in the primaries. I think I'll winnow that list down to the candidates who matter. I'll probably use a rule based on how many contributions received or the total dollar amount received.
* What is the amount contributed to Democrats and Republicans?
* Are contributor locations (by city or zip code) overrepresented?
* Gender might be an interesting variable to add, but it would be a pain to find and compare the names to some other public dataset (like voting records) in order to assign a gender. Even then, I couldn't guarantee that it's the same person and not another person with the same name.
* Which candidates received the most amount of contributions?
* How much money did candidates receive?

#### Add Political Party Variable
```{r}
assignParty <- function(x) {
    if(is.na(x)){
      NA
    }else if(x %in% c("Trump, Donald J.", "Cruz, Rafael Edward 'Ted'", "Walker, Scott", "Bush, Jeb",
                      "Rubio, Marco", "Kasich, John R.", "Graham, Lindsey O.", "Paul, Rand", 
                      "Fiorina, Carly", "Huckabee, Mike", "Carson, Benjamin S.", "Perry, James R. (Rick)",
                      "Jindal, Bobby", "Santorum, Richard J.", "Christie, Christopher J.") ){
      "R"
    }else if(x %in% c("Clinton, Hillary Rodham", "Sanders, Bernard", "O'Malley, Martin Joseph",
                      "Lessig, Lawrence")){
      "D"
    }else if(x == "Johnson, Gary"){
      "L"
    }else if(x == "Stein, Jill"){
      "G"
    }else if(x == "McMullin, Evan"){
      "I"
    }else {
        NA
    }
}

ne$cand_prty <- apply(ne['cand_nm'],1,assignParty)
ne$cand_prty <- as.factor(ne$cand_prty)
summary(ne$cand_prty)
```
```{r echo=FALSE}
ggplot(aes(x = cand_prty),
       data = ne) +
    geom_bar()
```

The vast majority of contributions were made to Democrats or Republicans. Of those two parties, about 50% more were made to Democrats.

```{r echo=FALSE, summary}
summary(ne)
```

The main features of interest are the candidates and the contribution amounts. Supportive features include contributor attributes (location, occupation, employer) and contribution dates.

Two election type contributions are classified as _Other_. I have no idea what that could be. 

#### Distribution of Contribution Amounts

The contribution amount variable has outliers on both sides of the data. On the left side, outliers are from refunds. On the right, outliers are from maximum allowed donations. Let's remove the [top and bottom 1%](http://r.789695.n4.nabble.com/removing-lower-and-upper-quantiles-from-an-arry-td3310479.html) and view.

```{r echo=FALSE}
ggplot(aes(x = contb_receipt_amt), 
       data = ne[ne$contb_receipt_amt >=
                       quantile(ne$contb_receipt_amt, .01) &
                       ne$contb_receipt_amt <=
                       quantile(ne$contb_receipt_amt, .99), ]) +
  geom_histogram(binwidth = 20)
```

It looks like most contributions are at the lower end of the scale, $250 or less. 

I've heard from past reading and news consumption that since Barack Obama's 2008 campaign, campaigns have seen a large uptick in the number of small individual contributions. In this case, most are below $250. Let's take a closer look at those contributions.

```{r echo=FALSE}
ggplot(aes(x = contb_receipt_amt), 
       data = ne) +
  geom_histogram(binwidth = 5) +
  xlim(-1, 250)
```

So, small amounts and easy denominations are popular, such as $25, $50, and $100. What about if we look at less than $100 contributions?

```{r echo=FALSE}
ggplot(aes(x = contb_receipt_amt), 
       data = ne) +
  geom_histogram(binwidth = 1) +
  xlim(-1, 100)
```

This looks like confirmation of the tendency to contribute in multiples of 5, like $5, $10, $15, etc. $10, $25, and $50 demoninations are the most popular.

#### Locations of Contributors

```{r echo=FALSE}
summary(ne$ne_city, maxsum = 25)
```

Most contributions come from Omaha and Lincoln. Some of these locations qualify as [villages](https://en.wikipedia.org/wiki/Village) and [hamlets](https://en.wikipedia.org/wiki/Hamlet_(place)). I wonder for some of these locations how the political contribution rate compares to that of a town or city. I won't delve into an answer here.

```{r echo=FALSE}
cities <- summary(ne$contbr_city)
(cities[1] + cities[2]) / nrow(ne)
```

Ignore the "OMAHA" label above - I couldn't find a way to remove it. Omaha and Lincoln account for about 57% of the contributions. Let's visualize this.

```{r echo=FALSE}
ggplot(aes(x = contbr_city), data = ne) +
  geom_bar()
```

Those top three cities certainly stand out. Omaha, Lincoln, Bellevue, then the rest. Let's [reduce](https://discussions.udacity.com/t/subsetting-candidates/231047) how many cities we're viewing to cities whose residents made over 500 contributions.

```{r echo=FALSE}
city_count = as.data.frame(table(ne$contbr_city))
ggplot(city_count[city_count$Freq > 500,], aes(x=Var1,y=Freq)) + 
    geom_bar(stat="identity")
```

The top three contributing cities also have the largest populations. Papillion and Bellevue are within Omaha's metro area. It seems like the number of contributions is a function of population.

#### Contribution Recipients

```{r echo=FALSE}
summary(ne$cand_nm, maxsum = 7)
```

It looks like at least three levels exist here. Clinton received over two times the number of contributions compared to Trump. Sanders fell between the two. Trump had about two times the number of contributions compared to Cruz and Carson. Cruz and Carson had at least seven times the number of contributions compared to the remaining candidates with Rubio coming the closest. 

```{r echo=FALSE}
cand <- summary(ne$cand_nm)
(cand['Clinton, Hillary Rodham'] + cand['Sanders, Bernard'])/nrow(ne)
```

Ignore the Clinton label above, I couldn't remove it. Clinton and Sanders accounted for about 61% of the total contributions. This is somewhat surprising for an overall conservative state, but recall that the metro areas accounted for about 57% of the contributions. Cities tend to be more liberal and vote Democrat, so this isn't surprising after all.

```{r echo=FALSE}
cand_count = as.data.frame(table(ne$cand_nm))
ggplot(cand_count[cand_count$Freq > 1000,], aes(x=Var1,y=Freq)) + 
    geom_bar(stat="identity")
```

The above chart is limited to only the candidates who received at least 1000 contributions. You get a good sense of the level of candidate support among the population of individual contributors in Nebraska.

# Univariate Analysis

_##### What is the structure of the dataset?_

The dataset contains mostly factor variables. Contribution amounts are numbers.

_##### What is/are the main feature(s) of interest in your dataset?_

The main primary features are the candidates and their received contributions.

_##### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?_

The secondary features are contributor attributes, like location (stored as a city or zipcode), occupation, employer, and contribution receipt date. The receipt date is probably the most noisy variable because it is unknown what the difference is between when the contributor made the contribution and when it was received, processed, and recorded. But it may prove an interesting if rough proxy for viewing spikes in contributions over time.

_##### Did you create any new variables from existing variables in the dataset?_

I created a variable for political party for each candidate as a supporting feature.

_##### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?_

I changed the receipt date from factor to date type so that I can possibly make a timeline later. The dataset was clean so I didn't do any tidying.

# Bivariate Exploration

#### Contributions Over Time

```{r echo=FALSE}
ggplot(aes(x = contb_receipt_dt, y = cand_nm), data = ne) +
  geom_jitter(alpha = .1, shape = 21, fill = 'blue') +
    scale_x_date(date_breaks = "6 months")
```

This reveals the number of contributions a candidate received over the two elections. Trump, Sanders, Cruz, Clinton, and Carson (with Rubio on the boundary), all had supporters who contributed often. This is a bit crowded though because some candidates hardly received contributions.

```{r}
ggplot(aes(x = contb_receipt_dt, y = cand_nm), 
       data = subset(ne, 
                     ne$cand_nm %in% unique(subset(data.frame(table(ne$cand_nm)), 
                                                       Freq > 100, select = Var1))$Var1)) +
    geom_jitter(alpha = .1, shape = 21, fill = 'blue') +
    scale_x_date(date_breaks = "6 months")
```

Here we see candidates who received over 100 contributions. This eliminates 14 candidates. Interestingly, by the density of the plot, it looks like Trump's contributions decreased as the general election progressed. I wonder if his supportors became less enthusiastic, supporters gave less often, or Trump didn't earn new supporters. 

```{r echo=FALSE}
ggplot(aes(x = contb_receipt_dt, y = cand_nm), 
       data = subset(ne, 
                     ne$cand_nm %in% unique(subset(data.frame(table(ne$cand_nm, 
                                                                        ne$contb_receipt_dt)),
                                                       Freq > 5, select = Var1))$Var1)) +
    geom_jitter(alpha = .1, shape = 21, fill = 'blue') +
    scale_x_date(date_breaks = "6 months")
```

This is a graph of candidates who received more than 5 contributions on some receipt date. I do not think this has much value. The receipt date is already a rough lagging proxy for support at some point in time. The data set did not indicate whether the receipt date is when the contributor made the contribution or when it was processed by the committee. Thus, this could say more about when the committee processed its received contributions than when a candidate received monetary support.

```{r}
ggplot(aes(x = contb_receipt_dt, y = cand_prty), data = ne) +
  geom_jitter(alpha = .1, shape = 21, fill = 'blue') +
    xlim("2015-06-01", "2016-12-01") +
    scale_x_date(date_breaks = "6 months")
```

Number of contributions to party candidates over time. It looks like the Republican candidate lost some support towards the end of the election. The Democrats had a mostly steady increase of contributions that sustained through 2016. The third party candidates garnered support during the general election. My hypothesis is that primary voters' ideal candidates did not win their respective primaries, so they migrated to these other candidates.

```{r}
ggplot(aes(x = contb_receipt_dt, y = cand_nm), data = ne) +
  geom_line() +
  scale_x_date(date_breaks = "6 months")
```

Here we can see the periods of time over which candidates received individual contributions. The standout information for me is the briefly considered candidates (like McMullin, Stein, and Christie), the three who received no to almost no support (Christie, Perry, Lessig, and Jindal), and the late arrivals (Stein, McMullin, and Johnson). As I noted above Stein and Johnson probably gathered support as constituents' favorites in their primaries lost.

```{r}
ne.conbt_amount_by_cand <- ne %>%
  filter(contb_receipt_amt > 0) %>% # remove negatives
  group_by(cand_nm, contb_receipt_dt) %>%
  summarize(mean_contb_amt = mean(contb_receipt_amt),
            median_contb_amt = median(contb_receipt_amt),
            n = n()) %>%
  ungroup() %>% 
  arrange(cand_nm)
head(ne.conbt_amount_by_cand)


ggplot(aes(x = contb_receipt_dt, y = median_contb_amt), data = ne.conbt_amount_by_cand) +
  geom_line(stat = 'summary', fun.y = median)
```

```{r}
# How can I plot the sum of contributions over time? (by candidate or party)
ne_short <- subset(ne, contb_receipt_amt > 0)
ggplot(aes(x = contb_receipt_dt, y = cumsum(contb_receipt_amt)), 
       data = ne_short) +
    geom_line()
```

```{r}
ggplot(aes(x= cand_nm, y = contb_receipt_amt), data = subset(ne, 
                     ne$cand_nm %in% unique(subset(data.frame(table(ne$cand_nm)), 
                                                       Freq > 250, select = Var1))$Var1)) +
    geom_boxplot() +
    ylim(0, 250)
```

A boxplot summary of contribution amounts by candidate. I eliminated the negative contributions, which I assume are returned contributions, and amounts over 250. This only includes candidates who received over 250 contributions. This clearly shows the typical range of contributions and that candidates received many outliers. Interestingly, the Republican candidates had higher interquartile ranges and medians compared to Democratic candidates. This could be an indicator that Republican supports are more affluent or more invested in their candidates.

# Bivariate Analysis

_##### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?_



_##### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?_

_##### What was the strongest relationship you found?_



# Multivariate Exploration

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}
ggplot(aes(x = contb_receipt_dt, y = cand_nm, color=cand_prty), data = ne) +
  geom_line() +
  scale_x_date(date_breaks = "6 months")
```
```{r}
ggplot(aes(x= cand_nm, y = contb_receipt_amt, color=cand_prty), data = subset(ne, 
                     ne$cand_nm %in% unique(subset(data.frame(table(ne$cand_nm)), 
                                                       Freq > 250, select = Var1))$Var1)) +
    geom_boxplot() +
    ylim(0, 250)
```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!

Resources
1. [row.names error](https://stackoverflow.com/questions/13239639/duplicate-row-names-error-reading-table-row-names-null-shifts-columns/22408965#22408965)